---
title: "'Time' out of Mind: Bob Dylan's Lyric Diction"
author: "Paul Reiners"
date: "7/20/2018"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, cache = FALSE, message = FALSE)
```

## Songs

As we can see, Bob Dylan was the most prolific in the 1960s and his song output
decreased over the decades.  However, in the 2000s, he made a comeback.

```{r pressure, echo=FALSE, message=FALSE}
# Load the libraries

#most of the libraries needed
library(dplyr) #data manipulation
library(ggplot2) #visualizations
library(gridExtra) #viewing multiple plots together
library(tidytext) #text mining
library(wordcloud2) #creative visualizations
library(stringr)

# Read in the data

bob_dylan <- read.csv("../../../data/bob_dylan_new.csv", stringsAsFactors = FALSE)
# Descriptive Statistics

#define some colors to use throughout
my_colors <- c("#E69F00", "#56B4E9", "#009E73", "#CC79A7", "#D55E00")

theme_lyrics <- function() 
{
    theme(plot.title = element_text(hjust = 0.5),
          axis.text.x = element_blank(), 
          axis.ticks = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          legend.position = "none")
}

bob_dylan %>%
    filter(decade != "NA") %>%
    group_by(decade) %>%
    summarise(number_of_songs = n()) %>%
    ggplot() + 
    geom_bar(aes(x = decade, y = number_of_songs), stat = "identity")  +
    theme(plot.title = element_text(hjust = 0.5),
          legend.title = element_blank(),
          panel.grid.minor = element_blank()) +
    ggtitle("Released Songs") +
    labs(x = NULL, y = "Song Count")
```

## Words

Unsurprising for a popular song writer, in general, but perhaps surprising for Bob
Dylan, in particular, "love" is the most frequently used (non-stop) word in
his lyrics.

```{r}
library(knitr) # for dynamic reporting
library(kableExtra) # create a nicely formated HTML table
library(formattable) # for the color_tile function

# Text Mining
# Tidy Text Format
# Data Formats and Tokenization
undesirable_words <- 
    c("rybody", "instrumental", "chorus", "babababababy", "bautiful")

# Word Frequency
full_word_count <- bob_dylan %>%
    unnest_tokens(word, lyrics) %>%
    group_by(song) %>%
    summarise(num_words = n()) %>%
    arrange(desc(num_words)) 

# Word Clouds
#unnest and remove stop, undesirable and short words
bob_dylan_words_filtered <- bob_dylan %>%
    unnest_tokens(word, lyrics) %>%
    anti_join(stop_words) %>%
    distinct() %>%
    filter(!word %in% undesirable_words) %>%
    filter(nchar(word) > 3)

bob_dylan_words_filtered %>%
    count(word, sort = TRUE) %>%
    top_n(10) %>%
    ungroup() %>%
    mutate(word = reorder(word, n)) %>%
    ggplot() +
    geom_col(aes(word, n), fill = my_colors[4]) +
    theme(legend.position = "none", 
          plot.title = element_text(hjust = 0.5),
          panel.grid.major = element_blank()) +
    xlab("") + 
    ylab("Song Count") +
    ggtitle("Most Frequently Used Words in Bob Dylan Lyrics") +
    coord_flip()

bob_dylan_words_counts <- bob_dylan_words_filtered %>%
    count(word, sort = TRUE) 

wordcloud2(bob_dylan_words_counts[1:300, ], size = .5)

# wordcloud2(bob_dylan_words_counts[1:300, ], figPath = "../img/glaser2.png", 
#            color = "random-dark", size = 1.5)
```

## Word Use Across the Decades

These results are perhaps the most interesting.  Although "love" is the most commonly
used word overall, "time" is the most frequently used word in Dylan's lyrics in
the 1960s, 1970s, and 1980s and is in the top 8 during the 1990s and 2000s.  (Although ["time" is a very commonly-used word](https://simple.wikipedia.org/wiki/Most_common_words_in_English) in general)  In the 2010s, neither "love" nor "time" is in the top 8
and things get a bit dark.

```{r words_by_decade}
timeless_words_decade <- bob_dylan_words_filtered %>% 
    filter(decade != 'NA') %>%
    group_by(decade) %>%
    count(word, decade, sort = TRUE) %>%
    slice(seq_len(8)) %>%
    ungroup() %>%
    arrange(decade,n) %>%
    mutate(row = row_number()) 

timeless_words_decade %>%
    ggplot(aes(row, n, fill = decade)) +
    geom_col(show.legend = NULL) +
    labs(x = NULL, y = "Song Count") +
    ggtitle("Timeless Words") + 
    theme_lyrics() +  
    facet_wrap(~decade, scales = "free", ncol = 5) +
    scale_x_continuous(  # This handles replacement of row 
        breaks = timeless_words_decade$row, # notice need to reuse data frame
        labels = timeless_words_decade$word) +
    coord_flip()
```

## Word Use Across Albums

Now we look at the most popular words for each album.

```{r words_by_album, fig.width=12, fig.height=16}
timeless_words_album <- bob_dylan_words_filtered %>% 
    filter(album != 'NA' & album != '') %>%
    group_by(album) %>%
    count(word, album, sort = TRUE) %>%
    slice(seq_len(8)) %>%
    ungroup() %>%
    arrange(album,n) %>%
    mutate(row = row_number()) 

album_year_df <- unique(bob_dylan[c("year", "album")])
timeless_words_album_year <- 
    merge(x = timeless_words_album, y = album_year_df, by = "album", all.x = TRUE)
timeless_words_album_year_sorted <- timeless_words_album_year[order(timeless_words_album_year$year),]

timeless_words_album_year_sorted$album2 <- 
    reorder(timeless_words_album_year_sorted$album, timeless_words_album_year_sorted$year)
timeless_words_album_year_sorted %>%
    ggplot(aes(row, n, fill = album)) +
    geom_col(show.legend = NULL) +
    labs(x = NULL, y = "Song Count") +
    ggtitle("Timeless Words by Album") + 
    theme_lyrics() +  
    facet_wrap(~album2, scales = "free", ncol = 4) +
    scale_x_continuous(  # This handles replacement of row 
        breaks = timeless_words_album$row, # notice need to reuse data frame
        labels = timeless_words_album$word) +
    coord_flip()
```

## Lexical Analysis

Lexical diversity has increased slightly over the years in Bob Dylan's lyrics.
In other words, he's been using a wider range of words.

```{r}
# Lexical Diversity
lex_diversity_per_year <- bob_dylan %>%
    filter(decade != "NA") %>%
    unnest_tokens(word, lyrics) %>%
    group_by(song,year) %>%
    summarise(lex_diversity = n_distinct(word)) %>%
    arrange(desc(lex_diversity)) 

diversity_plot <- lex_diversity_per_year %>%
    ggplot(aes(year, lex_diversity)) +
    geom_point(color = my_colors[3],
               alpha = .4, 
               size = 4, 
               position = "jitter") + 
    stat_smooth(color = "black", se = FALSE, method = "lm") +
    geom_smooth(aes(x = year, y = lex_diversity), se = FALSE,
                color = "blue", lwd = 2) +
    ggtitle("Lexical Diversity") +
    xlab("") + 
    ylab("") +
    scale_color_manual(values = my_colors) +
    theme_classic() + 
    theme_lyrics()

diversity_plot
```

Lexical density has decreased slightly over the years.  That is, his songs "say less" over time.
Given the high density and complexity of his lyrics from 1964--1966, this is probably not surprising.
However, lexical density has been making a comeback in his most recent songs.

```{r}
# Lexical Density
lex_density_per_year <- bob_dylan %>%
    filter(decade != "NA") %>%
    unnest_tokens(word, lyrics) %>%
    group_by(song,year) %>%
    summarise(lex_density = n_distinct(word)/n()) %>%
    arrange(desc(lex_density))

density_plot <- lex_density_per_year %>%
    ggplot(aes(year, lex_density)) + 
    geom_point(color = my_colors[4],
               alpha = .4, 
               size = 4, 
               position = "jitter") + 
    stat_smooth(color = "black", 
                se = FALSE, 
                method = "lm") +
    geom_smooth(aes(x = year, y = lex_density), 
                se = FALSE,
                color = "blue", 
                lwd = 2) +
    ggtitle("Lexical Density") + 
    xlab("") + 
    ylab("") +
    scale_color_manual(values = my_colors) +
    theme_classic() + 
    theme_lyrics()

density_plot
```

## TF-IDF

See [Wikipedia](https://en.wikipedia.org/wiki/Tf–idf) for a definition of TF-IDF.
Using TF-IDF on albums is quite informative.  We see which words are important
on a particular album.
 
```{r fig.width=12, fig.height=16}
tfidf_words_album <- bob_dylan %>%
    unnest_tokens(word, lyrics) %>%
    distinct() %>%
    filter(!word %in% undesirable_words & album != '') %>%
    filter(nchar(word) > 3) %>%
    count(album, word, sort = TRUE) %>%
    ungroup() %>%
    bind_tf_idf(word, album, n) %>%
    arrange(desc(tf_idf))

top_tfidf_words_album <- tfidf_words_album %>% 
    group_by(album) %>% 
    slice(seq_len(8)) %>%
    ungroup() %>%
    arrange(album, tf_idf) %>%
    mutate(row = row_number())

top_tfidf_words_album %>%
    ggplot(aes(x = row, tf_idf, fill = album)) +
    geom_col(show.legend = NULL) +
    labs(x = NULL, y = "TF-IDF") + 
    ggtitle("Important Words using TF-IDF by Album") +
    theme_lyrics() +  
    facet_wrap(~album, 
               ncol = 4, 
               scales = "free") +
    scale_x_continuous(  # this handles replacement of row 
        breaks = top_tfidf_words_album$row, # notice need to reuse data frame
        labels = top_tfidf_words_album$word) +
    coord_flip()
```

## Acknowledgements

This article is highly indebted to the techniques used in ["Lyric Analysis with NLP & Machine Learning with R"](https://www.datacamp.com/community/tutorials/R-nlp-machine-learning) written
by [Debbie Liske](https://www.datacamp.com/profile/debbieliskefa5156febfc047aeb79dee6c802d3d04).

## Code

My code is available in the [bob-dylan-haikus](https://github.com/paul-reiners/bob-dylan-haikus) project.
