---
title: "Bob Dylan Lyric Sentiment Analysis"
author: "Paul Reiners"
date: "7/27/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r initialization, echo=FALSE, include=FALSE}
library(dplyr) #Data manipulation (also included in the tidyverse package)
library(tidytext) #Text mining
library(tidyr) #Spread, separate, unite, text mining (also included in the tidyverse package)
library(widyr) #Use for pairwise correlation

#Visualizations!
library(ggplot2) #Visualizations (also included in the tidyverse package)
library(ggrepel) #`geom_label_repel`
library(gridExtra) #`grid.arrange()` for multi-graphs
library(knitr) #Create nicely formatted output tables
library(kableExtra) #Create nicely formatted output tables
library(formattable) #For the color_tile function
library(circlize) #Visualizations - chord diagram
library(memery) #Memes - images with plots
library(magick) #Memes - images with plots (image_read)
library(yarrr)  #Pirate plot
library(radarchart) #Visualizations
library(igraph) #ngram network diagrams
library(ggraph) #ngram network diagrams

#Define some colors to use throughout
my_colors <- c("#362A28", "#6F3036", "#4C3735", "#725047", "#8B675E", "#A5817B")

#Customize ggplot2's default theme settings
#This tutorial doesn't actually pass any parameters, but you may use it again in future tutorials so it's nice to have the options
theme_lyrics <- function(aticks = element_blank(),
                         pgminor = element_blank(),
                         lt = element_blank(),
                         lp = "none")
{
  theme(plot.title = element_text(hjust = 0.5), #Center the title
        axis.ticks = aticks, #Set axis ticks to on or off
        panel.grid.minor = pgminor, #Turn the minor grid lines on or off
        legend.title = lt, #Turn the legend title on or off
        legend.position = lp) #Turn the legend on or off
}

#Customize the text tables for consistency using HTML formatting
my_kable_styling <- function(dat, caption) {
  kable(dat, "html", escape = FALSE, caption = caption) %>%
    kable_styling(bootstrap_options = c("striped", "condensed", "bordered"),
                  full_width = FALSE)
}

### Sushi Data
dylan_data <- read.csv('../../../data/bob_dylan_new.csv', stringsAsFactors = FALSE, row.names = 1)

glimpse(dylan_data) #Transposed version of `print()`

### Good Clean Fun: prince_tidy
undesirable_words <- 
  c("rybody", "instrumental", "chorus", "babababababy", "bautiful", "don")
#Create tidy text format: Unnested, Unsummarized, -Undesirables, Stop and Short words
dylan_tidy <- dylan_data %>%
  unnest_tokens(word, lyrics) %>% #Break the lyrics into individual words
  filter(!word %in% undesirable_words) %>% #Remove undesirables
  filter(!nchar(word) < 3) %>% #Words like "ah" or "oo" used in music
  anti_join(stop_words) #Data provided by the tidytext package

glimpse(dylan_tidy) #From `dplyr`, better than `str()`.
```

In this article, we apply the data science technique of sentiment analysis to
the lyrics of Bob Dylan.

## Lexical diversity over time

We can see Bob Dylan's lexical diversity has increased over time.

```{r pressure, echo=FALSE}
word_summary <- dylan_tidy %>%
    filter(decade != "NA") %>%
  mutate(decade = ifelse(is.na(decade),"NONE", decade)) %>%
  group_by(decade, song) %>%
  mutate(word_count = n_distinct(word)) %>%
  select(song, Released = decade, word_count) %>%
  distinct() %>% #To obtain one record per song
  ungroup()

boxplot(word_count ~ Released, #Formula
           data = word_summary, #Data frame
           main = "Lexical Diversity Per Decade") 
```

## Song Count Per Year

```{r song_count, echo=FALSE}
songs_year <- dylan_data %>%
  select(song, year) %>%
  filter(!is.na(year)) %>%
  group_by(year) %>%
  summarise(song_count = n())

id <- seq_len(nrow(songs_year))
songs_year <- cbind(songs_year, id)
label_data = songs_year
```

Bob Dylan was the most prolific in the mid-1960s and the mid-1970s (with a trough in-between).
No surprise there.  There is also a surprise peak in 2009.  This is because both <i>Together Through Life</i>
and <i>Christmas in the Heart</i> came out that year.  (Arguably, one might want to
exclude <i>Christmas in the Heart</i>, but the sentiments were presumably his even
if he didn't write the words.)

```{r, echo=FALSE}
ggplot(songs_year, aes(x = year, y = song_count)) +
  geom_bar(stat = "identity", fill = alpha("darkred", 0.7)) +
  scale_x_continuous("year", breaks = songs_year$year) + 
  theme(axis.text.x = element_text(angle=90)) + 
  ylab("song count")
```

## Detailed Sentiment Analysis
```{r echo=FALSE, message=FALSE}
dylan_bing <- dylan_tidy %>%
  inner_join(get_sentiments("bing"))

dylan_nrc <- dylan_tidy %>%
  inner_join(get_sentiments("nrc"))

dylan_nrc_sub <- dylan_tidy %>%
  inner_join(get_sentiments("nrc")) %>%
  filter(!sentiment %in% c("positive", "negative"))
```

### Overall Sentiment

We use two well-known sentiment lexicons: 

 * [NRC Word-Emotion Association Lexicon](https://saifmohammad.com/WebPages/NRC-Emotion-Lexicon.htm)
 * [Bing sentiment lexicon](https://www.searchenginejournal.com/sentiment-analysis/235945/)
 
Each has their advantages and disadvantages.  Their results are sometimes inconsistent.
Also, it's arguable whether one should be doing automated sentiment analysis on poetry
(or song lyrics).  That sort of seems to be missing the point, somehow.  Nevertheless,
let's look at some of the results.
 
```{r echo=FALSE}
nrc_plot <- dylan_nrc %>%
  group_by(sentiment) %>%
  summarise(word_count = n()) %>%
  ungroup() %>%
  mutate(sentiment = reorder(sentiment, word_count)) %>%
  #Use `fill = -word_count` to make the larger bars darker
  ggplot(aes(sentiment, word_count, fill = -word_count)) +
  geom_col() +
  guides(fill = FALSE) + #Turn off the legend
  theme_lyrics() +
  labs(x = NULL, y = "Word Count") +
  scale_y_continuous(limits = c(0, 5000)) + #Hard code the axis limit
  ggtitle("Bob Dylan NRC Sentiment") +
  coord_flip()

plot(nrc_plot)

bing_plot <- dylan_bing %>%
  group_by(sentiment) %>%
  summarise(word_count = n()) %>%
  ungroup() %>%
  mutate(sentiment = reorder(sentiment, word_count)) %>%
  ggplot(aes(sentiment, word_count, fill = sentiment)) +
  geom_col() +
  guides(fill = FALSE) +
  theme_lyrics() +
  labs(x = NULL, y = "Word Count") +
  scale_y_continuous(limits = c(0, 4000)) +
  ggtitle("Bob Dylan Bing Sentiment") +
  coord_flip()

plot(bing_plot)
```

### Song sentiments over the years

I think it's very interesting that Bob Dylan's lyrics peaked in positivity in 1980.
This was exactly in the middle of his "gospel period".

```{r echo=FALSE, warning=FALSE}
dylan_polarity_year <- dylan_bing %>%
  count(sentiment, year) %>%
  spread(sentiment, n, fill = 0) %>%
  mutate(polarity = positive - negative,
         percent_positive = positive / (positive + negative) * 100)

polarity_over_time <- dylan_polarity_year %>%
  ggplot(aes(year, polarity, color = ifelse(polarity >= 0,my_colors[5],my_colors[4]))) +
  geom_col() +
  geom_smooth(method = "loess", se = FALSE) +
  geom_smooth(method = "lm", se = FALSE, aes(color = my_colors[1])) +
  theme_lyrics() + theme(plot.title = element_text(size = 11)) +
  xlab(NULL) + ylab(NULL) +
  ggtitle("Polarity Over Time")

relative_polarity_over_time <- dylan_polarity_year %>%
  ggplot(aes(year, percent_positive , color = ifelse(polarity >= 0,my_colors[5],my_colors[4]))) +
  geom_col() +
  geom_smooth(method = "loess", se = FALSE) +
  geom_smooth(method = "lm", se = FALSE, aes(color = my_colors[1])) +
  theme_lyrics() + theme(plot.title = element_text(size = 11)) +
  xlab(NULL) + ylab(NULL) +
  ggtitle("Percent Positive Over Time")

grid.arrange(polarity_over_time, relative_polarity_over_time, ncol = 2)
```

### Relationship Between Mood and Decade

Joy peaks in the 1980s (while quality of records dips) and sadness peaks in the 2010s.

```{r echo=FALSE, fig.height=12}
decade_mood <-  dylan_nrc %>%
  filter(decade != "NA" & !sentiment %in% c("positive", "negative")) %>%
  count(sentiment, decade) %>%
  group_by(decade, sentiment) %>%
  summarise(sentiment_sum = sum(n)) %>%
  ungroup()

decade_agg <- 
  setNames(aggregate(decade_mood$sentiment_sum, by=list(Category=decade_mood$decade), FUN=sum), c("decade", "sentiment_total"))
decade_mood_2 <- merge(decade_mood, decade_agg, by="decade")
decade_mood_2$sentiment_sum_norm <- decade_mood_2$sentiment_sum / decade_mood_2$sentiment_total
decade_mood <- decade_mood_2
 
## some a bar plot
par(mfrow = c(3, 2))

sixties = decade_mood[decade_mood$decade == "1960s", ]
seventies = decade_mood[decade_mood$decade == "1970s", ]
eighties = decade_mood[decade_mood$decade == "1980s", ]
nineties = decade_mood[decade_mood$decade == "1990s", ]
aughts = decade_mood[decade_mood$decade == "2000s", ]
tens = decade_mood[decade_mood$decade == "2010s", ]

barplot(sixties$sentiment_sum_norm, names.arg = sixties$sentiment, main = "decade=1960s", las=2)
barplot(seventies$sentiment_sum_norm, names.arg = seventies$sentiment, main = "decade=1970s", las=2)
barplot(eighties$sentiment_sum_norm, names.arg = eighties$sentiment, main = "decade=1980s", las=2)
barplot(nineties$sentiment_sum_norm, names.arg = nineties$sentiment, main = "decade=1990s", las=2)
barplot(aughts$sentiment_sum_norm, names.arg = aughts$sentiment, main = "decade=2000s", las=2)
barplot(tens$sentiment_sum_norm, names.arg = tens$sentiment, main = "decade=2010s", las=2)
```

### Real-Time Sentiment

Next we look at the relationship between the personal events in Bob Dylan's life
and the sentiments expressed in his songs at the time.  Perhaps surprising is that
his lyrics are the most positive during his "gospel period".  Perhaps less surprising is that they are also
positive during his time of domesticity during the late 1960s.

```{r echo=FALSE, warning=FALSE, message=FALSE}
events <- read.csv('../../../data/dylanEvents.csv', stringsAsFactors = FALSE)

year_polarity_bing <- dylan_bing %>%
  group_by(year, sentiment) %>%
  count(year, sentiment) %>%
  spread(sentiment, n) %>%
  mutate(polarity = positive - negative,
         ratio = polarity / (positive + negative)) #use polarity ratio in next graph

events %>%
  #Left join gets event years with no releases
  left_join(year_polarity_bing) %>%
  filter(event != " ") %>% #Account for bad data
  mutate(event = reorder(event, year), #Sort chart by desc year
         sentiment = ifelse(positive > negative,
                            "positive", "negative")) %>%
  ggplot(aes(event, polarity, fill = sentiment)) +
  geom_bar(stat = "identity") +
  theme_minimal() + theme(legend.position = "none") +
  xlab(NULL) +
  ggtitle("Sentiment by Events") +
  coord_flip()
```

### <i>Blood on the Tracks</i> Sentiment

Since <i>Blood on the Tracks</i> is perhaps Bob Dylan's most autobiographical album 
(either that or it's based on Chekhov short stories), let's analyze the sentiments 
in its lyrics.  Sadness is the predominant sentiment; not surprising for probably
the best 'break-up' record of all time.

```{r echo=FALSE, message=FALSE}
dylan_nrc %>%
    filter(album %in% "Blood On The Tracks") %>%
    group_by(sentiment) %>%
    summarise(word_count = n()) %>%
    ungroup() %>%
    mutate(sentiment = reorder(sentiment, word_count)) %>%
    ggplot(aes(sentiment, word_count, fill = -word_count)) +
    geom_col() +
    guides(fill = FALSE) +
    theme_minimal() + theme_lyrics() +
    labs(x = NULL, y = "Word Count") +
    ggtitle("Blood on the Tracks NRC Sentiment") +
    coord_flip()

dylan_tidy %>%
    filter(song %in% "Simple Twist Of Fate") %>%
    distinct(word) %>%
    inner_join(get_sentiments("nrc")) %>%
    ggplot(aes(x = word, fill = sentiment)) +
    facet_grid(~sentiment) +
    geom_bar() + #Create a bar for each word per sentiment
    theme_lyrics() +
    theme(panel.grid.major.x = element_blank(),
          axis.text.x = element_blank()) + #Place the words on the y-axis
    xlab(NULL) + ylab(NULL) +
    ggtitle("Blood On The Tracks Sentiment Words") +
    coord_flip()
```

## Acknowledgements

This article is highly indebted to the techniques described in ["Tidy Sentiment Analysis in R"](https://www.datacamp.com/community/tutorials/sentiment-analysis-R) written
by [Debbie Liske](https://www.datacamp.com/profile/debbieliskefa5156febfc047aeb79dee6c802d3d04).

## Code

My code is available in the [bob-dylan-haikus](https://github.com/paul-reiners/bob-dylan-haikus) project.
